Tests of measurement invariance via stochastic processes: {S}ome properties

### 2013-08-06 frontiers/frontiers.Rnw
fix the plot problem generated by sctest functions. didn't change anything in simall functions.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@230 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\SweaveOpts{engine = R, eps = FALSE, echo = TRUE, pdf=TRUE}
Alternatively, we could use sctest.lavaan function to get the results. 
<<Get Estimate Alternative Way>>=
DM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="DM", fplot=TRUE) 
CvM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="CvM", fplot=TRUE) 
#maxLM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, #stat="maxLM", fplot=TRUE) 
DM
CvM
#maxLM
@
\subsection{Ordinal Test}
However, in this dataset, $V$ is actually ordinal instead of continuous. There is only a partial ordering of individuals with respect to $V$,i.e. observations with the same level of $V$ have no unique ordering. The ordinal statistics proposed here are similar to those described in Equation and above, except that we focus on "bins" of individuals at each level of the ordinal variable. That is, instead of aggregating over all $i=1,...n$ individuals, we first compute cumulative proportions $t_{l}(l=1,...,m-1)$ associated with the first $m-1$ levels of $V$. We then aggregate the cumulative scores only over $i_{l}=\lfloor n \cdot t_{l} \rfloor$. Test statistics related can be written as 
\begin{eqnarray}
    \label{eq:wdm}
      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~ \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1/2}\max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
    \label{eq:maxlmo}
      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~\left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}\sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
\end{eqnarray}

@@

<<Get Estimate Alternative Way for Ordinal>>=
WDMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="WDMo", fplot=TRUE) 
#maxLMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, #stat="maxLMo", fplot=TRUE)
WDMo
#maxLMo
@

@@

\section{Simulation 1}

In Simulation 1, we examined the extent to which the proposed tests
are sensitive to the specific model parameter that violates measurement
invariance.  If, say, a factor loading violates measurement
invariance, it is plausible that this violation impacts other
parameters such as factor covariances or other loadings 
associated with the factor in question.  Thus, the goal of the
simulation was to examine the extent to which the proposed tests
attribute the measurement invariance violation to the parameters that
are truly in violation.

\subsection{Methods}
To examine these specificity issues, we generated data from a
two-factor model with three indicators each (the same model used in
Merkle and Zeileis, \citeyearNP{MerZei13}). The measurement invariance
violation occurred in one of three parameters: the factor loading
associated with the first indicator $\lambda_{11}$, the unique variance associated
with the first indicator $\epsilon_{11}$, or the factor covariance $\sigma_{12}$.  
We then tested for measurement invariance in seven subsets of parameters: the three
individual parameters noted above, all six factor loadings $\lambda_{11}--\lambda_{62}$, 
all six unique variances $\epsilon_{11}--\epsilon_{66}$. The path diagram is shown in  
Figure~\ref{fig:path}.

\begin{figure}
\caption{Path diagram representing the factor analysis model 
         used for simulations.}
\label{fig:path}
\includegraphics{PathDiagram}
\end{figure}

Power and Type I error were examined across three sample sizes
($n=100,200,500$) and 17 magnitudes of invariance violations. These
violations were based on the auxiliary variable $V$: individuals below
the 50th percentile of $V$ deviated from individuals above the 50th
percentile  by $d$ times the parameters' asymptotic standard errors
(scaled by $\sqrt n$), with $d=0,0,25,0.5,...,4$. 
For each combination of sample size ($n$) $\times$ violation
magnitude ($d$) $\times$ violating parameter, 5000 datasets were
generated and tested.

\subsection{Results}
Full simulation results are presented in Figure~\ref{fig:sim11res} to
Figure~\ref{fig:sim13res}; all three test statistics exhibited very
similar results, so the figures include only results for the CvM statistic.
Figure~\ref{fig:sim11res} displays power curves as a function of 
violation magnitude for the first
factor loading $\lambda_{11}$, with panels reflecting the parameters 
being tested and lines reflecting $n$.
Figures~\ref{fig:sim12res} and Figure~\ref{fig:sim13res} display 
similar power curves when the factor covariance 
$\phi_{12}$ and error variance $\epsilon_{11}$ violate measurement
invariance, respectively.

From these figures, one generally observe that the tests isolate 
the parameter violating measurement invariance.  Additionally, the tests 
have somewhat-higher power to detect measurement invariance violations 
in the factor loading and factor covariance parameters, as opposed to the error variance
parameter.  Finally, simultaneous tests of all factor loadings or all error
parameters result in decreased power, as compared to the situation
where one tests only the violating parameter.  This occurs because, in
testing a subset of parameters (only one of which violates measurement
invariance), we are effectively dampening the signal of a measurement
invariance violation.

In summary, we found that the proposed tests can attribute
measurement invariance violations to the correct parameter.  This
provides evidence that, in practice, one can have confidence in the
tests' abilities to locate the measurement invariance violation.  Of
course, this statement is qualified by the fact that, in this
simulation, the model was correctly specified.  In the following
simulations, we examine the tests' performance when the model is
misspecified.

%% Example of how to call a simulation in R:
<<sim1>>=
## seed for replication
RNGkind(kind = "default", normal.kind = "default")
set.seed(1090)

if(file.exists("sim1.rda")) {
  load("sim1.rda")
} else {
  ## Arguments lead to small-scale simulation:
  sim1 <- simulation(nobs=c(100,200,500), diff=seq(0,4,.25), parms=c("loading","var","error"))
  save(sim1, file="sim1.rda")
}

## Add to column names so that plots look nicer
sim1$nobs <- paste("n=", sim1$nobs, sep="")

#Not sure how to express all loadings and all error terms in greek letters.
greekExprList<-expression(lambda[11],epsilon[11],sigma[12],lambda[11]--lambda[62],
                          epsilon[11]--epsilon[66])
greekpar<-as.expression(greekExprList)

@

%% Example of how to create a figure of results
\begin{figure}
\caption{Simulated power curves for Simulation 1.Violation Parameter is loading}
\label{fig:sim11res}
\setkeys{Gin}{width=\textwidth}
<<sim11res, fig=TRUE, height=7, width=7>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey1 <- simpleKey(unique(sim1$nobs), points = TRUE, lines = TRUE)
print(xyplot(power ~ diff | pars, group = ~ nobs, data = sim1,
             subset = diff <= 4 & parms == "loading" & test=="CvM" &  pars %in% 
             c("1","7","13","1:6","7:12"),
              strip = function(which.given, which.panel, var.name,
                        strip.levels = FALSE,
                        strip.names = TRUE, ...) {
       	      strip.default(which.given, which.panel,
       	               var.name = greekpar[which.panel],
                       strip.levels = FALSE,
                       strip.names = TRUE, ...)},
             layout = c(3,2),type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey1,
             as.table=TRUE))
@
\end{figure}

%% Example of how to create a figure of results
\begin{figure}
\caption{Simulated power curves for Simulation 1.Violation Parameter is covariance}
\label{fig:sim12res}
\setkeys{Gin}{width=\textwidth}
<<sim12res, fig=TRUE, height=7, width=7>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
print(xyplot(power ~ diff | pars, group = ~ nobs, data = sim1,
             subset = diff <= 4 & parms == "var" & test == "CvM" &  pars %in% 
             c("1","7","13","1:6","7:12"),
              strip = function(which.given, which.panel, var.name,
                        strip.levels = FALSE,
                        strip.names = TRUE, ...) {
       	      strip.default(which.given, which.panel,
       	               var.name = greekpar[which.panel],
                       strip.levels = FALSE,
                       strip.names = TRUE, ...)},
             layout = c(3,2),
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey1,
             as.table=TRUE))
@
\end{figure}

\begin{figure}
\caption{Simulated power curves for Simulation 1.Violation Parameter is error terms}
\label{fig:sim13res}
\setkeys{Gin}{width=\textwidth}
<<sim13res, fig=TRUE, height=7, width=7>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
print(xyplot(power ~ diff | pars, group = ~ nobs, data = sim1,
             subset = diff <= 4 & parms == "error" & test == "CvM" &  pars %in%
             c("1","7","13","1:6","7:12"),
              strip = function(which.given, which.panel, var.name,
                        strip.levels = FALSE,
                        strip.names = TRUE, ...) {
       	      strip.default(which.given, which.panel,
       	               var.name = greekpar[which.panel],
                       strip.levels = FALSE,
                       strip.names = TRUE, ...)},
             layout = c(3,2),
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey1,
             as.table=TRUE))
@
\end{figure}

%###########################################################################################
\section{Simulation 2}
In Simulation 2, we examine the extent to which the results of
Simulation 1 are robust to model misspecification. Specifically, we
generate data from the factor analysis model used in the previous
section, but with adding an extra loading from the second factor to the first
indicator (of the first factor).  The estimated model lacks this extra
loading, however. The goal of this simulation is to examine the proposed statitics'
power and spicificity under model misspecification.

\subsection{Method}
Measurement invariance violations could occur in
each of the three parameters from Simulation 1 (factor loading,factor covariance,unique
variance), and they could also occur in the extra,unmodeled loading. 
Sample size and magnitude of measurementinvariance violation were manipulated 
in the same way as they were in Simulation 1.The tested parameters are the same as Simulation
1,too.

\subsection{Results}
Results are presented by the paramter that violated measurement invariance.
Figure~\ref{fig:sim21res} displays results when the unmodeled loading violates 
measurement invariance. One can generally observe that 
the first loading and error variance parameter showed high measurement invariance 
violation simultaneously, while factor covariance did not demonstrate violation.

Figure~\ref{fig:sim22res} displays results when the loading $\lambda_{11}$ violates
measurement invariance. Even though lacking one loading associated with the first indicator, 
CvM statistic showed enough power to detect the first loading violation specifically. 
Similar pattern of specificity and power were observed when the violation parameter 
was $\sigma_{12}$ (Figure~\ref{fig:sim23res}) and $\epsilon_{11}$ (Figure~\ref{fig:sim24res}).

In summary, the proposed test statistics appear robust to unmodeled loading parameters,so long
as the umodeled parameter does not violate measurment invariance. If the unmodeled parameter 
dose violate measurement invariance, however, then the tests assign this violation to other 
parameters that do not violate measurement invariance. The impacted parameters include the error
variance and other loadings associated with the manifest variables that had the unmodeled 
loading. Thus, prior to testing measurement invariance, it is important to study the extent to
which the hypothesized model includes all parameters of importance.

%% Example of how to call a simulation in R:
<<sim2>>=
## seed for replication
RNGkind(kind = "default", normal.kind = "default")
set.seed(1090)

if(file.exists("sim2.rda")) {
  load("sim2.rda")
} else {
  ## Arguments lead to small-scale simulation:
  sim2 <- simulation(nobs=c(100,200,500), diff=seq(0,4,.25), parms=c("extra",
                    "extra+loading", "extra+var","extra+error"))
  save(sim2, file="sim2.rda")
}

## Add to column names so that plots look nicer
sim2$nobs <- paste("n=", sim2$nobs, sep="")

@

%% Example of how to create a figure of results
\begin{figure}
\caption{Simulated power curves for Simulation 2.Violation Parameter is extra loading}
\label{fig:sim21res}
\setkeys{Gin}{width=\textwidth}
<<sim21res, fig=TRUE, height=7, width=7>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
mykey2 <- simpleKey(unique(sim2$nobs), points = TRUE, lines = TRUE)
print(xyplot(power ~ diff |pars , group = ~ nobs, data = sim2,
             subset = diff <= 4 & parms == "extra" & test == "CvM" & pars %in%
              c("1","7","13","1:6","7:12"),
              strip = function(which.given, which.panel, var.name,
                        strip.levels = FALSE,
                        strip.names = TRUE, ...) {
       	      strip.default(which.given, which.panel,
       	               var.name = greekpar[which.panel],
                       strip.levels = FALSE,
                       strip.names = TRUE, ...)},
             layout = c(3,2),
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey2))
@
\end{figure}

%% Example of how to create a figure of results
\begin{figure}
\caption{Simulated power curves for Simulation 2.Violation Parameter is loading, 
         but model misspecified}
\label{fig:sim22res}
\setkeys{Gin}{width=\textwidth}
<<sim22res, fig=TRUE, height=7, width=7>>=
print(xyplot(power ~ diff | pars, group = ~ nobs , data = sim2,
             subset = diff <= 4 & parms == "extra+loading" & test == "CvM" & pars %in% 
             c("1","7","13","1:6","7:12"),
              strip = function(which.given, which.panel, var.name,
                        strip.levels = FALSE,
                        strip.names = TRUE, ...) {
       	      strip.default(which.given, which.panel,
       	               var.name = greekpar[which.panel],
                       strip.levels = FALSE,
                       strip.names = TRUE, ...)},
             layout = c(3,2),
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey2))
@
\end{figure}

%% Example of how to create a figure of results
\begin{figure}
\caption{Simulated power curves for Simulation 2.Violation Parameter is var, 
         but model misspecified}
\label{fig:sim23res}
\setkeys{Gin}{width=\textwidth}
<<sim23res, fig=TRUE, height=7, width=7>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
print(xyplot(power ~ diff | pars, group = ~ nobs , data = sim2,
             subset = diff <=4 & parms == "extra+var" & test == "CvM" & pars %in%
             c("1","7","13","1:6","7:12"),
              strip = function(which.given, which.panel, var.name,
                        strip.levels = FALSE,
                        strip.names = TRUE, ...) {
       	      strip.default(which.given, which.panel,
       	               var.name = greekpar[which.panel],
                       strip.levels = FALSE,
                       strip.names = TRUE, ...)},
             layout = c(3,2),
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey2))
@
\end{figure}

%% Example of how to create a figure of results
\begin{figure}
\caption{Simulated power curves for Simulation 2.Violation Parameter is var, 
         but model misspecified}
\label{fig:sim24res}
\setkeys{Gin}{width=\textwidth}
<<sim24res, fig=TRUE, height=7, width=7>>=
trellis.par.set(theme = canonical.theme(color = FALSE))
## 
print(xyplot(power ~ diff | pars, group = ~ nobs , data = sim2,
             subset = diff <=4 & parms == "extra+error" & test == "CvM" & pars %in% 
             c("1","7","13","1:6","7:12"),
              strip = function(which.given, which.panel, var.name,
                        strip.levels = FALSE,
                        strip.names = TRUE, ...) {
       	      strip.default(which.given, which.panel,
       	               var.name = greekpar[which.panel],
                       strip.levels = FALSE,
                       strip.names = TRUE, ...)},
             layout = c(3,2),
             type = "b", xlab="Violation Magnitude", ylab="Power", key=mykey2))
@
\end{figure}

>>>>

\SweaveOpts{engine = R, eps = TRUE, echo = TRUE, pdf=TRUE}
Alternatively, we could use sctest.lavaan function to get the results.

<<Get Estimate Alternative Way>>=
DM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="DM", fplot=TRUE) 
DM
CvM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="CvM", fplot=TRUE) 
CvM
#maxLM<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, #stat="maxLM", fplot=TRUE) 
#maxLM
@

\begin{figure}
\label{fig:statistics}
\includegraphics[page=1]{Rplots}
\end{figure}

\begin{figure}
\label{fig:statistics}
\includegraphics[page=2]{Rplots}
\end{figure}

\subsection{Ordinal Test}
However, in this dataset, $V$ is actually ordinal instead of continuous. There is only a partial ordering of individuals with respect to $V$,i.e. observations with the same level of $V$ have no unique ordering. The ordinal statistics proposed here are similar to those described in Equation and above, except that we focus on "bins" of individuals at each level of the ordinal variable. That is, instead of aggregating over all $i=1,...n$ individuals, we first compute cumulative proportions $t_{l}(l=1,...,m-1)$ associated with the first $m-1$ levels of $V$. We then aggregate the cumulative scores only over $i_{l}=\lfloor n \cdot t_{l} \rfloor$. Test statistics related can be written as 
\begin{eqnarray}
    \label{eq:wdm}
      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~ \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1/2}\max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
    \label{eq:maxlmo}
      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~\left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) \right\}^{-1}\sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
\end{eqnarray}

@@

<<Get Estimate Alternative Way for Ordinal>>=
WDMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="WDMo", fplot=TRUE)
WDMo 
maxLMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="maxLMo", fplot=TRUE)
maxLMo
@

\begin{figure}
\label{fig:statistics}
\includegraphics[page=3]{Rplots}
\end{figure}

\begin{figure}
\label{fig:statistics}
\includegraphics[page=4]{Rplots}
\end{figure}

@@

\section{Simulation}
